---
title: "Batch analysis across many cohorts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Batch analysis across many cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

A common workflow is to run the same NMF profiling on many independent
expression matrices -- different cohorts, subgroups, tissues, or conditions --
and then compare the factors found across them. `run_nmf_batch()` drives
[`NMFprofileR()`](../reference/NMFprofileR.html) over a named list of cohorts and
returns a single consolidated summary tagged by cohort, so the results can be
pooled and traced back to their source.

The driver is:

* **resilient** -- one cohort failing does not abort the batch (`on_error = "continue"`);
* **resumable** -- cohorts whose results already exist are skipped and loaded from disk (`skip_existing = TRUE`);
* **self-describing** -- every summary row carries the cohort's `Run_ID`.

```{r setup}
library(NMFprofileR)
data("example_expression_data")
dim(example_expression_data)
```

## Defining cohorts

A cohort is just an expression matrix in the form
[`NMFprofileR()`](../reference/NMFprofileR.html) accepts (genes in rows with
gene-symbol rownames, samples in columns). `cohorts` is a **named** list; the
names become both the `run_id` and the output subdirectory for each cohort.

Here we split the bundled example data into two illustrative cohorts:

```{r cohorts}
half <- ncol(example_expression_data) %/% 2
cohorts <- list(
  cohort_A = example_expression_data[, seq_len(half)],
  cohort_B = example_expression_data[, (half + 1):ncol(example_expression_data)]
)
vapply(cohorts, ncol, integer(1))
```

In practice the cohorts might come from splitting one matrix by a grouping
variable, or from reading several matrices off disk into a named list.

## Running the batch

Every argument other than the ones the driver manages
(`expression_data`, `output_prefix`, `run_id`) is passed straight through to
[`NMFprofileR()`](../reference/NMFprofileR.html). Each cohort's outputs land in
`output_dir/<name>_Results/`, and the consolidated summary is also written to
`output_dir/Batch_Consolidated_Summary.tsv`.

```{r run, eval = FALSE}
batch <- run_nmf_batch(
  cohorts,
  output_dir        = file.path(tempdir(), "batch_demo"),
  nmf_rank          = 2:4,
  nmf_nrun          = 20,
  gprofiler_sources = c("GO:BP", "REAC"),
  enrichment_cache  = file.path(tempdir(), "gprofiler_cache")
)

batch$consolidated   # one row per factor, per rank, per cohort (tagged Run_ID)
batch$skipped        # cohorts skipped because results already existed
batch$failures       # cohorts that errored (Cohort, Reason)
```

The consolidated table has the same per-factor columns as a single run, with a
leading `Run_ID`:

```{r illustrate, echo = FALSE}
data.frame(
  Run_ID          = c("cohort_A", "cohort_A", "cohort_B"),
  Rank            = c(2L, 2L, 2L),
  Factor          = c("Factor_1", "Factor_2", "Factor_1"),
  Num_Samples     = c(14L, 12L, 15L),
  Percent_Samples = c(53.8, 46.2, 57.7)
)
```

## Resuming an interrupted batch

Because `skip_existing = TRUE` by default, re-running the same call after an
interruption only computes the cohorts that have not finished yet; the completed
ones are read back from their saved `nmf_profile` bundle (or
`Consolidated_Summary.tsv`) and still included in `batch$consolidated`. Set
`skip_existing = FALSE` to force every cohort to re-run.

## Making a batch fast and robust

* **Cache enrichment.** Pass `enrichment_cache = <dir>` so repeated g:Profiler
  queries (common when cohorts share genes, or when re-running) are served from
  disk instead of the network. Queries are keyed by a hash of their genes,
  background, sources, and settings.
* **Parallelize the fits.** Pass `nmf_parallel = TRUE` (optionally with
  `nmf_cores`) to spread each cohort's consensus runs across a local cluster.
  For a fixed `nmf_seed` the parallel path returns the same factorization as the
  sequential one.
* **Keep going on failure.** With `on_error = "continue"` (the default) a cohort
  that errors is recorded in `batch$failures` and the batch proceeds; use
  `on_error = "stop"` to fail fast.
* **Skip the second enrichment pass.** Marker-gene enrichment is opt-in
  (`emit_marker_genes = FALSE` by default), which halves the number of
  g:Profiler queries per cohort.

## Notes

The heavy steps depend on the `NMF` package and on live g:Profiler network
calls, so the batch chunk above is shown but not executed; the cohort-building
chunks run for real.
```{r sessioninfo, eval = FALSE}
utils::sessionInfo()
```