---
title: "Getting started with NMFprofileR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with NMFprofileR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**NMFprofileR** runs a full multi-rank consensus Non-negative Matrix
Factorization (NMF) over a transcriptomics matrix and layers g:Profiler
functional enrichment on top. For each rank *k* it fits consensus NMF, assigns
genes and samples to factors, runs enrichment, computes silhouette widths, and
(optionally) writes a tree of tables and figures to disk.

The heavy steps depend on the `NMF` package and on live g:Profiler network
calls, so those chunks below are shown but not executed. The lightweight steps
run for real.

```{r setup}
library(NMFprofileR)
data("example_expression_data")
dim(example_expression_data)
example_expression_data[1:5, 1:3]
```

## The one-call pipeline

`NMFprofileR()` orchestrates the whole workflow. It writes results under
`output_prefix` **and** returns an `nmf_profile` object with everything in
memory.

```{r run, eval = FALSE}
result <- NMFprofileR(
  expression_data = example_expression_data,
  nmf_rank        = 2:5,
  output_prefix   = file.path(tempdir(), "NMF_demo"),
  nmf_nrun        = 20,
  gprofiler_sources = c("GO:BP", "GO:MF", "GO:CC", "REAC", "TF")
)
result
#> <nmf_profile>
#>   Ranks fitted : 2, 3, 4, 5
#>   Samples      : 53
#>   Runtime      : 3.2 mins
#>   g:Profiler   : e111_eg58_p18_...
#>   Output dir   : .../NMF_demo_Results
```

The result keeps the fitted objects, gene assignments, and enrichment tables so
you never have to re-read files:

```{r access, eval = FALSE}
summary(result)              # consolidated per-factor metrics across ranks
result$fits[["3"]]           # fitted NMF object for k = 3
result$basis_genes[["3"]]    # every gene -> its dominant factor (argmax)
result$marker_genes[["3"]]   # factor-specific marker genes (specificity-scored)
result$sample_assignments[["3"]]
result$enrichment$combined[["3"]]
```

Pass `write_files = FALSE` to compute entirely in memory without creating any
files -- useful for scripting, examples, and tests.

## Preprocessing

The first stage validates and filters the matrix. It requires gene-symbol
rownames, resolves duplicated symbols (`on_duplicate_genes`, default
`"collapse_max"`), drops all-zero genes, and selects highly-variable genes.

```{r preprocess}
m <- nmf_preprocess(
  example_expression_data,
  expression_threshold = 10,
  variance_quantile    = 0.9
)
dim(m)
```

`variance_scale = "log"` ranks variability on `log2(x + 1)` instead of the raw
scale, which reduces the mean-variance confound of count data. The matrix fed
to NMF is always on the original non-negative scale.

## Composable stages

Every stage is exported and works in memory, so you can build a custom workflow
or inspect intermediate results.

```{r stages, eval = FALSE}
fit     <- nmf_fit(m, rank = 3, nrun = 20)
genes   <- nmf_basis_genes(fit)          # argmax assignment (all genes)
markers <- nmf_marker_genes(fit)         # specificity-scored markers (subset)
samples <- nmf_sample_assignments(fit)   # sample -> factor + silhouette

gene_sets <- split(genes$Gene, genes$Factor)
enr <- nmf_enrichment(gene_sets, background_genes = rownames(m))
```

## Choosing the number of components

`nmf_rank_diagnostics()` lays the rank-selection metrics side by side. In a real
run you would pass `result$rank_metrics`; here is a small illustrative table:

```{r diagnostics}
rank_metrics <- data.frame(
  rank       = 2:5,
  cophenetic = c(0.98, 0.95, 0.90, 0.86),
  dispersion = c(0.90, 0.82, 0.75, 0.70),
  silhouette = c(0.61, 0.55, 0.48, 0.44)
)
nmf_rank_diagnostics(rank_metrics)
```

A common heuristic is to pick the largest rank before the cophenetic
correlation drops sharply, cross-checked against a silhouette peak.

## Reproducibility

Every run records a manifest (parameters, package version, runtime), the exact
g:Profiler database version behind the enrichment, and full session information,
and the UMAP embedding is seeded -- so a result can be traced to the inputs and
data snapshot that produced it (`result$provenance`).

## Installation note

NMFprofileR depends on the Bioconductor packages `ComplexHeatmap` and
`circlize`. When installing from GitHub, enable the Bioconductor repositories
first:

```{r install, eval = FALSE}
setRepositories(ind = c(1, 2))   # CRAN + Bioconductor
devtools::install_github("MilesGithub/NMFprofileR")
```